betweenCell.Var

### 2020-05-08 Chap3_MLH1var.rmd
starting work on last ms outline

---
title: 'Evolution of Sex-Specific Recombination Rate'
date: '`r Sys.Date()`'
author: "April L. Peterson, Bret Payseur"
output:
  word_document:
    toc: yes
    toc_depth: '5'
  pdf_document:
    toc: yes
    toc_depth: '5'
  html_document:
    toc: yes
    toc_depth: '5'
    toc_float: yes
always_allow_html: yes
bibliography: refs_Main_Report.bib
---

```{r set.up, echo=FALSE, include=FALSE}

@@

#Intro

#Material Methods Draft v1

### Mouse Husbandry

To access natural genetic variation for *Mus musculus*, wild derived inbred strains were used. The _Mus musculus_ strains PWD/PhJ, PERC/Eij, WSB/EiJ, LEWE/EiJ, MSM/MsJ, MOLF/EiJ, CAST/EiJ, CZECHII/EiJ were purchased from Jackson labs (Maine, USA). The strains of KAZ/TUA, TOM/TUA, AST/TUA, and HMI/TUA were generated from frozen embryos purchased  from Biological Resource Center (BRC) at Riken (Ibaraki, Japan).  (https://en.brc.riken.jp) and rederived by the UW-Biotech center (Madison UW).
The related murid species *Mus caroli* CAROLI/EiJ and *Mus spretus* SPRET/EiJ were purchased from Jackson labs (https://www.jax.org) and *Mus spicilegus* SPIC/Eij from Riken. We use abbreviated  superscript notation for the remainder of the paper (e.g. _musculus^WSB^_). All mice were housed UW-Madison Biotech and MSC facilities following RARC approved protocols.  A breeding colony of wild derived _Mus musculus_ mice sampled from Gough Island (G) is maintained at UW Veterinary school facilities.

### Tissue Collection and Immunohistochemistry

The same dry-down spread technique was applied to both spermatocytes and oocytes based on [@peters_1997], with adjustment for volumes. Spermatocyte spreads were collected and prepared as described in [@peterson2019]. The majority of mice used MLH1 quantification were from mice between 5 and 12 weeks, while juvenile mice aged between 12 to 15 days were used for DMC1 quantification. Both pairs of ovaries were collected from embryo or neonate mice between 16 to 21 embryonic days and 0 to 48 hours old respectively. Whole testes and decapsulated ovaries were incubated in 3ml or 300ul of hypotonic solution for 45min respectively. Fifteen microliters of cell slurry (masticated gonads) were transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The following morning, slides were treated with a Photoflow wash (Kodak, cite). Slides were stored at -20*C if not stained immediately.

To visualize the structure of meiotic chromosomes we used antibody markers for the centromere (CREST) and lateral element of the synaptonemal complex (SC) (SYCP3). Double strand breaks (DSB) and COs were visualized with DMC1 and MLH1 respectively. The staining protocol was based on that in [@anderson1999] and [@koehler2002]. Antibodies and slide blocking was performed in 1X antibody dilution buffer (ADB) (normal donkey serum (Jackson ImmnuoResearch), 1X PBS, bovine serum albumin (Sigma), Triton X-100 (Sigma) ).

Following a 30 minute blocking wash in ABD, each slide was incubated with 60ul of a primary antibody master mix for 48 hours at 37* C. The master mix recipe contained polyclonal anti rabbit anti-MLH1 (Calbiochem; diluted 1:50) or anti rabbit anti-DMC1) (mix of DMC1), anti goat polyclonal anti-SYCP3, (Abcam; diluted 1:50), and anti-human polyclonal antibody to CREST (Antibodies, Inc; diluted 1:200) suspended in ADB. Slides were washed twice in 50ml ADB before the first round of secondary antibody incubation for 12 hours at 37 degrees. Alexa Fluor 488 donkey anti-rabbit IgG (Invitrgoen, location; diluted to 1:100) and Coumarin AMCA donkey anti-human IgG (Jackson ImmunoResearch; diluted to 1:200)  were suspended in ADB. The last incubation of Alexa Fluor 568 donkey anti-goat (Invitrogen; diluted 1:100) was incubated at 1:100 for 2 hours at 37 degrees. Slides were fixed with Prolong Gold Antifade (Invitrogen) for 24 hours after a final wash in 1x PBS.

### Image Processing

Images were capture using a Zeiss Axioplan 2 microscope with AxioLab camera and Axio Vision software (Zeiss, Cambridge, UK). Preprocessing, including cropping, noise reduction, and histogram adjustments, was performed using Photoshop (v13.0). Image file names were anonymized before manual scoring of MLH1 or DMC1 foci using photoshop.

### Analysis

To estimate the numbers of percursor sites which may lead to crossovers, the mean **DMC1** foci per cell for a single mouse from a subset of srains *musculus^PWD^*, *musculus^MSM^*, *domesticus^WSB^*, and *domesticus^G^* was quantified. SC morphology and CREST/centromere foci number was used to stage spermatocytes, in early and late zygotema stages.

**MLH1** foci were quantified from cells with intact and complete karyotypes (19 acrocentric bivalents and XY for spermatocytes or 20 acrocentric bivalents for oocytes) and distinct MLH1 foci. A quality score, between 1 to 5, was assigned to each cell based on staining quality and general spread of bivalents. With scores of 1 and 2 having the highest quality. Cells and coded as a score of 5 were excluded from the final analysis. The distributions of MLH1 counts per cell was assessed for normality (supplemental figure). MLH1 foci located on the XY in spermatocytes were excluded from this study.

**M1 Mixed Model** $$mouse \ av.\ dependant\ variable\ ~=~ subsp * sex + rand(strain) + \varepsilon $$

**M2 Linear Model** $$mouse \ av.\ dependant\ variable\ ~=~ subsp * sex * strain + \varepsilon$$

**M3 Linear Model** $$mouse \ av.\ dependant\ variable\ ~=~ sex * strain + \varepsilon$$

**M4 Linear Model** $$sex\ specific\ mouse \ av.\ dependant\ variable\ ~=~ subsp * strain + \varepsilon$$

**M5 Linear Model** $$sex\ specific\ mouse \ av.\ dependant\ variable\ ~=~  strain + \varepsilon$$

A series of mixed models and general linear models were used as an analetical framework. We built the full mixed model (**M1**) to interpret the patterns of variation using the lmer() from the lmer4 package (cite) in R (v3.5.2)[@Rstudio]. Strain was coded as a random effect and tested using exactRLRT() to reflect a random sample of natural genetic variation from each subspecies. The main effects, subspecies, strain, and interaction were coded as fixed and tested by a chi-square test comparing the full and reduced models using drop1() and anova().

This model allows us to estimate and test the fixed effects (sex and subsp), and variation in mean MLH1 counts due to strain effect. We use the subspecies term as a way to quantify between group variation (or divergence) and the random strain effect to quantify within group variation (or polymorphism) in a sex specific manner. The remaining general linear models, (**M2-M3**), were used to follow up on results from the full mixed model. The general linear models, **M4** were used to analyze sex specific data sets.

Our goal was to design an analytical framework which accounts for the nested nature across categories across all the individual mice (e.g. sex, strain subspecies) and can be applied to multiple (meiotic traits) dependent variables. In addition to the MLH1 count variables for individual mice, mean MLH1 count per cell, variance and coefficient of variance for MLH1 count per cell, we analyzed SC based features: mouse average of total SC per cell, mouse average of single SC length, mouse mean IFD^norm^ and IFD^raw^ for double crossover chromosomes, and the average position of crossover for single crossover bivalents.

These traits were choosen because they are markers of phenomena in the recombinaiton pathway and provide more detailed information about the recombination landscape. In order to examine crossover interference, inter-focal distance (IFD) was measured from the single bivalents containing two crossovers. Normalized measures were calculated by the position divided by the length of the SC for that single bivalent.

### Single bivalent characterization

To characterize the SC length of individual chromosomes, two image analysis algorithms were used to quantify SC area for individual cells [@wang_2019_sc] and for individual bivalents [@peterson2019]. Both apply a 'skeletonizing' transformation to the synapsed chromosomes which produces a single pixel wide 'trace' of the original chromosome shape. Hand measures of pactyene chromosomes were performed by using ImageJ/Fiji (v1.52) [@fiji].

The **total SC** area per cell was quantified from the pachtyene cell images. To remove erroneous SC isolation, outliers were visually assessed at the mouse level and removed from the data set. Mouse means were calculated from cell-wide total SC lengths in `r formatC( length(  (MLH1.merge.Skel %>% filter(subsp != "Cast")  )$Random.Name), format="d", big.mark=",")` out of `r formatC( length(  (MLH1_data %>% filter(subsp != "Cast")  )$Random.Name), format="d", big.mark=",")` cells with MLH1 counts.

The **DNA CrossOver** algorithm isolates single straightened bivalents shapes from an image, and returns SC length, and location of green and blue signal, reflecting MLH1 foci and cetromere signal respectively. This algorithm substantially speeds the accurate measurement of bivalents, but has the limitation that not all bivalents per cell can be isolated due to overlapping bivalents. In this data set, isolation rates per cell (the number of all bivalents per cell) range from `r min(algorithm.stat.category$mean_pass.rate)` ( _molossinus^MSM^_ male) to `r max(algorithm.stat.category$mean_pass.rate)` ( _musculus^KAZ^_ female).

A curation step was applied to the total single bivalent data set to remove poor measures from the algorithm. From the total set of cell images, `r formatC( length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName ), format="d", big.mark=",")` bivalent objects were isolated by the image analysis software. After a human curation step , `r formatC( length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID ), format="d", big.mark=",")` single-bivalent observations remained. The accuracy of the algorithm is high compared to hand measures after this curation step [@peterson2019]. From the curated single bivalent data set the **proportions** of bivalents by crossover number were quantified and tested by chi-square tests with the prop.test().

To account for confounding effects of sex chromosomes from pooled samples of bivalents, we considered a **reduced data** set including only bivalents with SC lengths below the 2nd quartile in cells with at least 17 of 20 single bivalent measures. The 'short bivalent' data set included the four or five shortest bivalents, and excluded the X bivalent in oocytes. A total of `r sum(Q1.real.short_mouse_table$n_obs)` 'short' bivalents were isolated from `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="female") )$ncells)` oocytes and `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="male") )$ncells)` spermatocytes. Although this smaller data set has decreased power, it offers a more comparable set of single bivalents to compare between the sexes.  A 'long bivalent' data set was calculated as above the 4th quartille in SC lengths per cell.

# Results

```{r FOR.PLOT.organizing.data, echo=FALSE, warning=FALSE}
#set the order of another column, based on another variable. consistant order for ploting

#apply this before ploting
MLH1_data <- MLH1_data %>%
  arrange(strain, sex, category, mouse) %>%
  mutate(Original.Name = factor(Original.Name)) #another category that you want the order to match
MLH1_data <- with(MLH1_data, MLH1_data[order(sex, strain),])

# sort your dataframe, by the focal categories
MLH1_data$mouse <- factor(MLH1_data$mouse, levels=unique(MLH1_data$mouse))
#use the above line for when order is out of wak

#MLH1_data <- with(MLH1_data, MLH1_data[order(category),])
AP_mouse_table.HQ$var <- as.numeric(AP_mouse_table.HQ$var)

AP_mouse_table.HQ_target <- AP_mouse_table.HQ[!(is.na(AP_mouse_table.HQ$sex) | AP_mouse_table.HQ$sex==""), ]

remove_strains <- c("TOM","AST","CZECH","CAST","HMI","SPRET","CAROLI", "F1", "PERC")
AP_mouse_table.HQ_target <- AP_mouse_table.HQ_target[ ! AP_mouse_table.HQ_target$strain %in% remove_strains, ]

var.plot <- ggplot(AP_mouse_table.HQ_target, aes(y=var, x = sex, color=sex))+geom_jitter()+facet_wrap(~strain)+ggtitle("Plots of mouse Variances for MLH1 counts")+
  scale_color_manual(values=c("royalblue", "limegreen"))+theme_bw()

```{r MLH1.MIXED.MODEL.setup, echo=FALSE, warning=FALSE, eval=TRUE}
#DF.HetC.MixedModel.HQ
#this blck is duplicated in setupData.rm

#https://rcompanion.org/handbook/G_03.html
#testing random effects within lme

DF.HetC.MixedModel <- AP_mouse_table_w.Ages[AP_mouse_table_w.Ages$species == "M.musculus",]

#AP_mouse_table excluide Qual5 cells
DF.HetC.MixedModel.HQ <- AP_mouse_table.MLH1_HQ
DF.HetC.MixedModel.HQ <- add_species(DF.HetC.MixedModel.HQ)

#main df with variables coded as chr
DF.HetC.MixedModel.HQ$subsp <- factor(DF.HetC.MixedModel.HQ$subsp, ordered = FALSE)
DF.HetC.MixedModel.HQ$sex <- factor(DF.HetC.MixedModel.HQ$sex, ordered = FALSE)
DF.HetC.MixedModel.HQ$strain <- factor(DF.HetC.MixedModel.HQ$strain, ordered = FALSE)
#DF.HetC.MixedModel.HQ$species <- factor(DF.HetC.MixedModel.HQ$species, ordered = FALSE)

#kable(table(DF.HetC.MixedModel.HQ$sex, DF.HetC.MixedModel.HQ$strain))

all.cell.MM.HQ <- ddply(.data=DF.HetC.MixedModel.HQ,
                 .(sex, strain),
                 summarize,
                 all.cells = sum(Ncells)
)

```{r table.for.plot, echo=FALSE, warning=FALSE}
#why do I add in a different table?
#this move this to the setup file
#MLH1_data  %>% filter(quality < 5)  -- low quality cells removed

#pooled
strain.table.HQ_Fig.og <- ddply(MLH1_data  %>% filter(quality < 5), c("subsp","strain", "sex"), summarise,
                        Nmice =length(unique(mouse)),
                        Ncells  =length(nMLH1.foci),
                        mean_co =as.numeric(format(round(mean(nMLH1.foci), 3 ), nsmall=3) ),
                        cV = cv(nMLH1.foci, na.rm = TRUE),
                        var = as.numeric(format(round(   var(nMLH1.foci, na.rm = TRUE),3), nsmall=3)), 
                        sd   = round(sd(nMLH1.foci, na.rm = TRUE), 3),
                       #Standard error (mice vs cells)
                        se  = round(sd / sqrt(Ncells), 3)

                       #range_mouse_means = range()
                   #consider adding crhm proportions?
    )

strain.table.HQ_Fig <-  strain.table.HQ_Fig.og %>% filter(strain != "F1" & strain != "PERC" &
                       strain != "TOM" &  strain != "AST" & strain != "CZECH" & strain != "CAROLI" & strain != "HMI")

#adjust order of plot here
strain.table.HQ_Fig$strain <- factor(strain.table.HQ_Fig$strain,levels =c(
     "MSM", "MOLF",   
  "PWD", "SKIVE",  "KAZ",
  "WSB","G","LEW","PERC",
                     "CAST",
                         "SPRET","SPIC" ), order=T )

```{r Fig1.top, echo=FALSE, warning=FALSE, message=FALSE}

Fig1_v1_strain <- ggplot(strain.table.HQ_Fig, aes(y=mean_co, x=strain, color=strain, shape=sex))+ 
  ylab("Average MLH1 Foci per Cell")+

  geom_point(aes(shape = sex), size=3, position=position_dodge(.5))+

    geom_errorbar(aes(ymin=mean_co-(se*2), ymax=mean_co+(se*2) ), width=.3, position = position_dodge(.5), size=1)+

  ylim(c(18, 35))+
 scale_color_manual(values=colors_of_strains)+

    theme(legend.position="none",
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line=element_line(colour = "black"),
  axis.title.y=element_blank(),
  axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank())+
#remove y labfor male

   #segments
  annotate("segment", x=.75, xend=4.5, y=19.5, yend=19.5, colour="black", size=1)+
  #Musc
  annotate("segment", x= 5.5, xend=10.2, y=19.5, yend=19.5, colour="black", size=1)
  #mol
#  annotate("segment", x=11, xend=12.25, y=19.5, yend=19.5, colour="black", size=1)+
  #cast
#  annotate("segment", x=13, xend=14, y=19.5, yend=19.5, colour="black", size=1)+
  #outgroups
 # annotate("segment", x=14.5, xend=16, y=19.5, yend=19.5, colour="black", size=1)+
  #label

 #   annotate("text", x = c(2,4.4,6,8, 10), #list of all x points 
   #        y = c(18.5,18.5,33,18.5, 18.5),

  # label = c("M. m. domesticus","M. m. musculus", "M. m. molossinus", "M. m. castaneus", "Outgroups"),
    #       fontface = 'italic', hjust = 0.5, size=3.5)

```{r female.show.plots, echo=FALSE, fig.width = 4, fig.height=2.5, warning=FALSE, message=FALSE}

female_mouse.av <- subset(AP_mouse_table_w.Ages, select= -c(sex.y,sex.x,
                            subsp.y,subsp.x,
                    category.y,category.x, strain.x,
                    strain.y, mouse.1  )  )

female_mouse.av <- add_sex(female_mouse.av)
female_mouse.av <- add_strain(female_mouse.av)
female_mouse.av <- add_subsp(female_mouse.av)
female_mouse.av <- add_species(female_mouse.av)

female_mouse.avs_for.analysis <-  female_mouse.av %>% filter(sex== "female") %>% filter(species == "M.musculus")

female_mouse.avs_for.analysis$subsp <- factor(female_mouse.avs_for.analysis$subsp, ordered = FALSE)
female_mouse.avs_for.analysis$sex <- factor(female_mouse.avs_for.analysis$sex, ordered = FALSE)
female_mouse.avs_for.analysis$strain <- factor(female_mouse.avs_for.analysis$strain, ordered = FALSE)
female_mouse.avs_for.analysis$species <- factor(female_mouse.avs_for.analysis$species, ordered = FALSE)

Musc.female_mouse.av <-  female_mouse.avs_for.analysis %>% filter(sex== "female") %>% filter(subsp == "Musc")

female.specific.plot <- ggplot(female_mouse.avs_for.analysis, aes(x = as.factor(strain), y = mean_co, color= strain))+ geom_boxplot()+  geom_jitter(aes( alpha=0.5), width = .2 )+ ggtitle("")+scale_color_manual(values=colors_of_strains)+ylim(c(17,34))+
  theme(legend.position="none",

        axis.title.y =element_blank(), 
      axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(),
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line=element_line(colour = "black"))+geom_hline(yintercept = 20)

#+theme(axis.text.x = element_text(angle = 45, hjust = 1))

```{r male.show.plots, echo=FALSE, fig.width = 4, fig.height=2.5, warning=FALSE, message=FALSE}
#try customize the plot size

male.plot <- ggplot(just.male_mouse.av, aes(x = as.factor(strain), y = mean_co, color= strain))+geom_jitter( alpha=0.5, width = .2) +geom_boxplot()+ ggtitle("")+scale_color_manual(values=colors_of_strains)+ylim(c(17,34))+

   theme(legend.position="none", axis.text.x=element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(),  axis.title.y = element_blank(), 
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line=element_line(colour = "black"))+geom_hline(yintercept = 19)

#horizontal line = 20

```{r Figure1.Main.plot.three.plots, echo=FALSE, warning=FALSE, fig.width=7, fig.cap='*Figure 1 Mean MLH1 count distributions by strain.* A) Strain averages of MLH1 counts per cell. B) Female specific MLH1 count distributions. C)	Male specific MLH1 counts per cell.', message=FALSE}

Figure1.main.plot <- Fig1_v1_strain / (female.specific.plot + male.plot )

Figure1.main.plot + plot_annotation(
  title = '',
  subtitle = '',
  caption = ''
) + plot_annotation(tag_levels = 'A')

```{r rand.misc, echo=FALSE, include=FALSE, eval=FALSE}
#str(AP_mouse_table_w.Ages)

AP_mouse_table.HQ$var <- as.numeric(AP_mouse_table.HQ$var)

#remove sex.x == na, and strain.x=na or F1
AP_mouse_table.HQ <- AP_mouse_table.HQ[!(is.na(AP_mouse_table.HQ$sex) | AP_mouse_table.HQ$sex==""), ]

AP_mouse_table.HQ <- AP_mouse_table.HQ[!(is.na(AP_mouse_table.HQ$var) | AP_mouse_table.HQ$var==""), ]
#plot variance and mean (facet sex)

AP_mouse_table_w.Ages <- AP_mouse_table_w.Ages[!( AP_mouse_table_w.Ages$strain.x =="F1"), ]

VandMean.mouse <- ggplot(aes(x=mean_co, y = var), data=AP_mouse_table.HQ )+geom_point(aes(color=strain))+facet_wrap(~sex)+ggtitle("mouse MLH1 averages; variance by mean")

cVandMean.mouse <- ggplot(aes(x=mean_co, y = cV), data=AP_mouse_table.HQ )+geom_point(aes(color=strain))+facet_wrap(~sex)+ggtitle("mouse MLH1 averages; cV by means")

#str(AP_mouse_table.HQ)

cor(AP_mouse_table.HQ$mean_co[AP_mouse_table.HQ$sex == "female"], AP_mouse_table.HQ$var[AP_mouse_table.HQ$sex == "female"])
#[1] 0.5750399
cor(AP_mouse_table.HQ$mean_co[AP_mouse_table.HQ$sex == "male"], AP_mouse_table.HQ$var[AP_mouse_table.HQ$sex == "male"])
#0.321601

cor(AP_mouse_table.HQ$mean_co[AP_mouse_table.HQ$sex == "female"], AP_mouse_table.HQ$cV[AP_mouse_table.HQ$sex == "female"])
# 0.3515544
cor(AP_mouse_table.HQ$mean_co[AP_mouse_table.HQ$sex == "male"], AP_mouse_table.HQ$cV[AP_mouse_table.HQ$sex == "male"])
#-0.03713536

cor.test(AP_mouse_table.HQ$mean_co[AP_mouse_table.HQ$sex == "male"], 
    AP_mouse_table.HQ$var[AP_mouse_table.HQ$sex == "male"] )
# p-value = 8.623e-09

cor.test(AP_mouse_table.HQ$mean_co[AP_mouse_table.HQ$sex == "male"], 
    AP_mouse_table.HQ$cV[AP_mouse_table.HQ$sex == "male"] )

#remove mice with fewer than 10 cells

AP_mouse_table.HQ.10plus <- AP_mouse_table.HQ[AP_mouse_table.HQ$Ncells >= 10,]

VandMean.mouse.10plus <- ggplot(aes(x=mean_co, y = var), data=AP_mouse_table.HQ.10plus )+geom_point(aes(color=strain))+facet_wrap(~sex)+ggtitle("10 plus mouse MLH1 averages; variance by mean")

cor(AP_mouse_table.HQ.10plus$mean_co[AP_mouse_table.HQ.10plus$sex == "female"], AP_mouse_table.HQ.10plus$cV[AP_mouse_table.HQ.10plus$sex == "female"])

cor.test(AP_mouse_table.HQ.10plus$mean_co[AP_mouse_table.HQ.10plus$sex == "female"], 
    AP_mouse_table.HQ.10plus$var[AP_mouse_table.HQ.10plus$sex == "female"] )

#why did I male just a female table?
female.musc.cells <- MLH1_data[( MLH1_data$sex == "female") & (MLH1_data$subsp == "Musc" ), ]

female.musc.mice.table <- ddply(female.musc.cells, c("sex"), summarise,
                        Nmice = length(unique(mouse)),
                        Ncells  = length( nMLH1.foci),
                        mean_co = as.numeric(format(round(  mean( nMLH1.foci), 3 ), nsmall=3) ),
                        cV = cv( nMLH1.foci),
                        var = format(round(   var( nMLH1.foci),3), nsmall=3),
                        sd   = round(sd( nMLH1.foci), 3),
                        se   = round(sd / sqrt(Ncells), 3)
                        #quality?
)
```

## Genome-wide recombination rate estimates for both sexes

```{r MLH1.summ.stats, echo=FALSE}

#strain.table.HQ_Fig.og  all strains
#strain.table.HQ_Fig  extra strains removed

main.table <- strain.table.HQ_Fig.og  %>% filter(strain != "F1")

main.table <- add_subsp(main.table)
main.table <- add_species(main.table)

#main.table %>% filter(sex == "female")
#main.table %>% filter(sex == "male")

#G.HETC.val #1.2
#LEW.HETC.val  #1
#WSB.HETC.val  #1
#KAZ.HETC.val  #1.1

#SKIVE.HETC.val # 0.96
#PWD.HETC.val  #0.88
#MSM.HETC.val #0.9

#kable(head(main.table),  caption = "Table 1. MLH1 counts per cell summary statistics by strain")

#strain.table.HQ_Fig

#sum( (main.table  %>% filter(sex == "male"))$Ncells)
#the kable version is bulkier than the other versaion

```{r MLH1.sum.stats.Bret.edits, echo=FALSE}

AP_mouse_table.HQ <- add_species(AP_mouse_table.HQ)
#total mice
#length( unique(AP_mouse_table.HQ %>% filter(species == "M.musculus") )$mouse )

#average cell number for spermatocytes and oocytes
#mean(female_mouse.avs_for.analysis$Ncells, na.rm = T)
#mean( (male_mouse.av %>% filter(species == "M.musculus") )$Ncells, na.rm = T)
```

We used counts of MLH1 foci per cell to estimate genome-wide recombination rates in `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` wild-derived inbred strains sampled from three subspecies of house mice: *M. m. domesticus*, *M. m. musculus* and *M. m. molossinus*. Mean MLH1 foci counts for `r length( unique(AP_mouse_table.HQ_target)$mouse )` mice were quantified from an average of `r mean( (AP_mouse_table.HQ_target %>% filter(sex == "male") )$Ncells, na.rm = T)` spermatocytes per male (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "male"))$Ncells),  format="d", big.mark=",")` spermatocytes) and `r mean(female_mouse.avs_for.analysis$Ncells, na.rm = T)` oocytes per female (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells), format="d", big.mark=",")` oocytes).

Graphical comparisons between the two sexes reveal several patterns (Figure 1A). First, recombination rate differs between females and males in most strains. Second, the direction and magnitude of heterochiasmy varies among strains. Although the majority of strains show higher recombination rates in females (following the pattern in laboratory mice), two musculus strains and one molossinus strain exhibit male-biased heterochiasmy. Hence, relative recombination rates in the two sexes are evolutionarily labile.

The general patterns of heterochiasmy in house mouse are displayed in Figure 1A. Taking note of the direction and the magnitude of the sex differences, our results confirm two general patterns: i) genome-wide recombination rates averages are greater in females compared to males (female biased heterochiasmy) and ii) the degree of heterochiasmy (Female:Male ratio) is generally low, ranging from `r G.HETC.val` in _domesticus^G^_ to `r LEW.HETC.val` in _domesticus^LEW^_. Three notable exceptions of male biased heterochiasmy are the strains _musculus^PWD^_, _molossinus^MSM^_ and _musculus^SKIVE^_, with heterochiasmy values of `r PWD.HETC.val`, `r MSM.HETC.val` and `r SKIVE.HETC.val` respectively.

Separately examining the mouse means of MLH1 foci per cell for each sex points to distinct patterns of variation (Figure 1B-C). Female recombination rates are evenly distributed around the sex-wide mean of approximately 25 MLH1 foci per cell (Figure 1B). In stark contrast, males separate more clearly into two groups of strains with high rates (near 30 MLH1 foci per cell) and low rates (near 23 MLH1 foci per cell) (Figure 1C).

@@

                        #Standard error (mice vs cells)
                        mean.mouse_se  = mean(se, na.rm=TRUE)
    )
#F.strain.table

#F.strain.table

# I don't know why most of the table is blank
```

```{r mouse.var.plots.female, echo=FALSE, fig.width = 4, fig.height=2.5, include=FALSE, eval=FALSE}

@@

```{r male.strain.table, echo=FALSE}
M.strain.table <- ddply(fomo.male_mouse.av, c("strain"), summarise,
                        Nmice =length(unique(mouse)),

@@

#variance plots
musc.male.VAR.plot <- ggplot(Musc.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y =var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Musc male of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)

Mol.male.VAR.plot <- ggplot(Mol.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Mol male of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)

Dom.male.VAR.plot <- ggplot(Dom.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Dom male ofwithin mouse  MLH1 distributions")+scale_color_manual(values=colors_of_strains)

@@

```{r MAIN.TABLE1, echo=FALSE, eval=TRUE}
#write the full table to a supplemental table
#show the head (or shortened)
#kable(table(AP_mouse_table.HQ$category), col.names = c("Category", "mice"), label = "Number of mice by category" )
#kable(strain.table.HQ_Fig, "latex")

df.strain.table <- strain.table.HQ_Fig
#kable(df.strain.table)
#pander(df.strain.table)

#kable(strain.table.HQ_Fig, col.names = c("Subspecies", "Strain","Sex", "Number of Mice", "Number of Cells", "Mean MLH1 count per cell", "cV", "Var", "sd", "se"), caption = "Strain Level Summary Statistics" )  %>%
#  kable_styling() %>%
#  scroll_box(width = "800px", height = "200px")

#kable(table(MLH1_data$category), col.names = c("Category","Cells"), label = "Number of Cells by category" )

#write excel
#write.xlsx2(strain.table.HQ_Fig, "Main_MLH1_table.xlsx", sheetName = "Sheet1", 
#  col.names = TRUE, row.names = FALSE, append = FALSE)

## Partitioning variation in recombination rate

To situate variation in recombination rate within an evolutionary framework, we fit a series of models including subspecies, strain, and sex, to mean MLH1 foci counts from `r length(DF.HetC.MixedModel.HQ$mean_co)` mice. We began with a full mixed model (M1, see Methods), which showed that strain 
(random effect p < `r 1/1000000`), sex (p = `r MLH1.MMlmer_results$sex.results`), subspecies (p=`r MLH1.MMlmer_results$sub.sp`), and subspecies sex (p = `r MLH1.MMlmer_results$interaction.pval`) each significantly affect recombination rate.

After a general linear model including all factors as fixed effects (M2) revealed only weak contributions of subspecies, we focused on additional models designed to illuminate the role of strain and sex. A general linear model with these two variables (M3) identified two strains with particularly strong effects on recombination rate: _musculus^MSM^_ (p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]`)
and _domesticus^G^_ (p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]`). In addition, two strains exhibit strain-by-sex interactions:  _molossinus^MSM^_  (p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]`) and _musculus^PWD^_ (p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`).

We next fit general linear models separately for `r length(fomo.male_mouse.av$mouse)` males and `r length(female_mouse.avs_for.analysis$mouse)` females (M4; see methods). In the male dataset, three strains significantly affect recombination rate: _musculus^PWD^_ ((glm; p = `r sum_MLH1.male.glm.M4$coefficients[,4][8]`; effect = 
`r sum_MLH1.male.glm.M4$coefficients[8]` foci), and 
_musculus^SKIVE^_ (glm; p = `r sum_MLH1.male.glm.M4$coefficients[,4][10]`; effect = `r sum_MLH1.male.glm.M4$coefficients[10]`), and  _molossinus^MSM^_ (glm; p=`r sum_MLH1.male.glm.M4$coefficients[,4][9]`; effect  `r sum_MLH1.male.glm.M4$coefficients[9]`).

>>>>

---
title: "betweenCell.Var"
author: "April Peterson"
date: "5/5/2020"
output: html_document
---

load all the file set ups

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

@@

plot the variances

@@

                        #Standard error (mice vs cells)
                        mean.mouse_se  = mean(se, na.rm=TRUE)
    )

```{r mouse.var.plots.female, echo=FALSE, fig.width = 4, fig.height=2.5, include=FALSE, eval=TRUE}

@@

```{r male.strain.table, echo=FALSE}

M.strain.table <- ddply(fomo.male_mouse.av, c("strain"), summarise,
                        Nmice =length(unique(mouse)),

@@

#variance plots
musc.male.VAR.plot <- ggplot(Musc.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y =var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Musc male of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)
musc.male.VAR.plot

Mol.male.VAR.plot <- ggplot(Mol.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Mol male of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)
Mol.male.VAR.plot

Dom.male.VAR.plot <- ggplot(Dom.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Dom male ofwithin mouse  MLH1 distributions")+scale_color_manual(values=colors_of_strains)
Dom.male.VAR.plot
```

go over the main patterns (from the model framework)

Bret's suggestions

On the question of variance in MLH1 count, I agree that it is crucial to reduce experimental noise to find a signal.

I would suggest the following approach:
1.  Consider only high-quality cells.
2. Pool cells within a strain and within a sex and compute one estimate of the variance in MLH1 count for each sex in each strain.
3. Randomly permute cells among strains, keeping the structure of the dataset otherwise the same.
4. For each permutation, compute the among-strain variance in the MLH1 count variance (separately for each sex).
Compute the among-strain variance in the MLH1 count variance (separately for each sex) in your original data.
Compare your observed variance with the null distribution generated by permutations to compute a p-value (one for each sex).

```{r bret.filters, echo=FALSE}

@@

# Identifying mice with largest sample sizes

I have a hunch that using these mice might be the best (least risky way forward)
